### 2017-10-15 mcmc.tex
autocorrelation advice

Not being mathematicians, we don't know the precise definition of a
pdf\note{Seriously.  It has something to do with measure theory.}, but for our purposes here a pdf is
any single-valued (scalar) function that is non-negative everywhere
(the entire domain of $\pars$), and obeys a normalization condition
\begin{eqnarray}
0 &\leq& p(\pars) \quad \mbox{for all $\pars$}
\\ \label{eq:normp}
1 &=& \int p(\pars)\,\dd\pars
\quad ,
\end{eqnarray}
where, implicitly, the integral is over the full domain of $\pars$.
Importantly, although $p(\pars)$ is single-valued, $\pars$ can be
a multi-element vector or list or blob; it can be arbitrarily large
and complicated.
In data-analysis contexts, $\pars$ will often be the full blob of free
parameters in the model.
Implicitly, the integral in \equationname~(\ref{eq:normp}) is high-dimensional; it
has as many dimensions as there are elements or entries or components of $\pars$.
Also, if there are elements or entries or components of $\pars$ that are discrete
(that is, take on only integer values or equivalent),
then along those dimensions the integral becomes a discrete sum.
This latter is a detail to which we return below. HOGG DO WE?

@@

That is, if the expectation value $E[g]$ is computed by the sampling
approximation given in \equationname~(\ref{eq:expectationvalue}),
there is an uncertainty on that expectation given by
\begin{eqnarray}
  \Var[E_{p(\pars)}[g(\pars)]] &=& \frac{\tau_q}{K}\,\Var_{p(\pars)}[g(\pars)]
  \\
  &\approx& \frac{\tau_q}{K^2}\,\sum_{k=1}^K \left(g(\pars_k) -
    E_{p(\pars)}[g(\pars)]\right)^2
  \quad,
\label{eq:tauvar}
\end{eqnarray}
where we are being a bit loose with terminology because this is the
variance not on $E_{p(\pars)}[g(\pars)]$ but on our sampling-based
\emph{estimator} for
$E_{p(\pars)}[g(\pars)]$, and the $E_{p(\pars)}[g(\pars)]$ inside the sum is
that estimator too.
It is slightly off-topic here, but related to this, if you have a long
chain of length $K$ and with autocorrelation time $\tau$ and you thin
the chain down to every $\tau$th sample (instead of using all $K$
samples), you will get just as good sampling-based estimates as if you
had used all $K$ samples.

@@

Of course, the autocorrelation-time definition
(\ref{eq:autocorrelationtime}) does not by itself provide an
\emph{estimator} for the time.
We might first try the na\"ive estimator
\begin{eqnarray}\label{eq:tauestnaive}
\tau &\stackrel{?}{=}& 1 + 2\,\sum_{\Delta = 1}^{N} C(\Delta)
\end{eqnarray}
where the sum is over the autocorrelation function estimated based the chain
using \equationname~(\ref{eq:autocorrelationfn}).
In practice, this is a high variance estimator of the integrated
autocorrelation time and this variance doesn't go to zero for an infinitely
long chain.\note{%
  A.~Sokal has a clear discussion of this topic and several proposals for
  robust autocorrelation time estimation in a set of lecture notes
  \url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}.}
Instead, we can get a much more robust estimate of the autocorrelation time by
restricting the sum in \equationname~(\ref{eq:tauestnaive}) to $\Delta \le M$
where
$\tau \le M \ll N$
\begin{eqnarray}\label{eq:tauest}
\tau &=& 1 + 2\,\sum_{\Delta = 1}^{M} C(\Delta)
\end{eqnarray}
In practice, $M$ can be chosen iteratively to satisfy these constraints but it
will only be as reliable as the chain itself.
An estimate of $\tau$ using \equationname~(\ref{eq:tauest}) should only be
trusted if the chain is many times ($\sim 10$) longer than $\tau$.

@@

% *********** BJB talking:
% I suggest partitioning the discussion of different methods into the following
% two categories:
%
%  i) Methods for getting around the target distribution
% ii) Methods that change the target distribution to something
% other than the posterior pdf, to help the sampling somehow
%
% i) Would include things like M-H, the stretch move,
% Hamiltonian, slice sampling (maybe skip the latter)
% ii) would include things like annealing and Nested Sampling

%%%% Hogg talking:
% DFM: In this section, I think we should just summarize the other
% methods, not describe them in detail.  Then we can cite the relevant
% papers for the enthusiastic.

>>>>

Not being mathematicians, we don't know the precise definition of a
pdf\note{Seriously.  It has something to do with measure theory.}, but for our purposes here a pdf is
any single-valued (scalar) function that is non-negative everywhere
(the entire domain of $\pars$), and obeys a normalization condition
\begin{eqnarray}
0 &\leq& p(\pars) \quad \mbox{for all $\pars$}
\\ \label{eq:normp}
1 &=& \int p(\pars)\,\dd\pars
\quad ,
\end{eqnarray}
where, implicitly, the integral is over the full domain of $\pars$.
Importantly, although $p(\pars)$ is single-valued, $\pars$ can be
a multi-element vector or list or blob; it can be arbitrarily large
and complicated.
In data-analysis contexts, $\pars$ will often be the full blob of free
parameters in the model.
Implicitly, the integral in \equationname~(\ref{eq:normp}) is high-dimensional; it
has as many dimensions as there are elements or entries or components of $\pars$.
Also, if there are elements or entries or components of $\pars$ that are discrete
(that is, take on only integer values or equivalent),
then along those dimensions the integral becomes a discrete sum.
This latter is a detail to which we return below. \textbf{HOGG DO WE?}

@@

That is, if the expectation value $E[g]$ is computed by the sampling
approximation given in \equationname~(\ref{eq:expectationvalue}),
there is an uncertainty on that expectation given by
\begin{eqnarray}
  \Var[E_{p(\pars)}[g(\pars)]] &=& \frac{\tau_g}{K}\,\Var_{p(\pars)}[g(\pars)]
  \\
  &\approx& \frac{\tau_g}{K^2}\,\sum_{k=1}^K \left(g(\pars_k) -
    E_{p(\pars)}[g(\pars)]\right)^2
  \quad,
\label{eq:tauvar}
\end{eqnarray}
where we are being a bit loose with terminology because this is the
variance not on $E_{p(\pars)}[g(\pars)]$ but on our sampling-based
\emph{estimator} for
$E_{p(\pars)}[g(\pars)]$, and the $E_{p(\pars)}[g(\pars)]$ inside the sum is
that estimator too.
It is slightly off-topic here, but related to this, if you have a long
chain of length $K$ and with autocorrelation time $\tau$ and you thin
the chain down to every $\tau$th sample (instead of using all $K$
samples), you will get just as good sampling-based estimates as if you
had used all $K$ samples.

@@

Of course, the autocorrelation-time definition
(\ref{eq:autocorrelationtime}) does not by itself provide an
\emph{estimator} for the time. Estimating the auto-correlation time is
notoriously difficult: It is a two-point statistic, and two-point statistics
are much harder to estimate than one-point statistics.\note{See: All of cosmology!}
If you are not a hard-core user of MCMC, and if all you want is heuristic indicators
of convergence, then what you should take from this section is that the auto-correlation
time is involved in variance estimates through \equationname~(\ref{eq:tauvar}) but that
it is hard to estimate.

If you want to think about autocorrelation estimation:\note{One of us (DFM) has written
  much more about autocorrelation estimation in the \project{emcee} documentation
  at \url{http://emcee.readthedocs.io/en/develop/tutorials/autocorr/}.}
You might first try the na\"ive estimator
\begin{eqnarray}\label{eq:tauestnaive}
\tau &\stackrel{?}{=}& 1 + 2\,\sum_{\Delta = 1}^{N} C(\Delta)
\end{eqnarray}
where the sum is over the autocorrelation function estimated based the chain
using \equationname~(\ref{eq:autocorrelationfn}).
In practice, this is a high variance estimator of the integrated
autocorrelation time and this variance doesn't go to zero for an infinitely
long chain.\note{%
  A.~Sokal has a clear discussion of this topic and several proposals for
  robust autocorrelation time estimation in a set of lecture notes
  \url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}.}
Instead, we can get a much more robust estimate of the autocorrelation time by
restricting the sum in \equationname~(\ref{eq:tauestnaive}) to $\Delta \le M$
where
$\tau \le M \ll N$
\begin{eqnarray}\label{eq:tauest}
\tau &=& 1 + 2\,\sum_{\Delta = 1}^{M} C(\Delta)
\end{eqnarray}
In practice, $M$ can be chosen iteratively to satisfy these constraints but it
will only be as reliable as the chain itself.
An estimate of $\tau$ using \equationname~(\ref{eq:tauest}) should only be
trusted if the chain is many times ($\sim 100$) longer than $\tau$.\footnote{This
  point is also emphasized at the aforementioned tutorial
  \url{http://emcee.readthedocs.io/en/develop/tutorials/autocorr/}, where
  a number like 50 is recommended.}

@@
